---
title: "**brglm2**: Reduced-bias inference in generalized linear models"
author:  |
  | Dr Ioannis Kosmidis
  | Department of Statistical Science
  | University College London
date: 29 June 2016
output:
  revealjs::revealjs_presentation:
    theme: night
    center: true
    reveal_options:
      slideNumber: true
      previewLinks: true
    transition: default
    background_transition: default
    fig_caption: true
    fig.width: 7
    fig.height: 5
bibliography: brglm2.bib
css: mycss.css
---


```{r options, echo = FALSE, tidy = TRUE}
library("knitr")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "\n", x) }
knit_hooks$set(document = hook2)
````

# **brglm** (with no "2")

## Binomial-response GLMs
* Bias reduction for binomial-response generalized linear models through adjusted score functions [@firth:93; @kosmidisdot:09]
* Iterative maximum likelihood fits on appropriately adjusted binomial responses and totals [see @kosmidisdot:14]

---

```{r brglm_c1, echo = TRUE, comment = '', prompt = TRUE, tidy = FALSE}
library("brglm")
brglm(cbind(grahami, opalinus) ~ height + diameter + light + time,
      data = lizards, family = binomial("logit"))
```

## Other GLMs
```{r brglm_c2, echo = TRUE, error = TRUE, comment = '', prompt = TRUE, tidy = TRUE}
## A Gamma example, from McCullagh & Nelder (1989, pp. 300-2)
clotting <- data.frame(u = c(5,10,15,20,30,40,60,80,100, 5,10,15,20,30,40,60,80,100),
                       conc = c(118,58,42,35,27,25,21,19,18,69,35,26,21,18,16,13,12,12),
                       lot = factor(c(rep(1, 9), rep(2, 9))))
glm(conc ~ lot*log(u), data = clotting, family = Gamma)
```

<!-- starts a new slide -->
---

```{r brglm_c3, echo = TRUE, error = TRUE, comment = '', tidy = TRUE}
## A Gamma example, from McCullagh & Nelder (1989, pp. 300-2)
brglm(conc ~ lot*log(u), data = clotting, family = Gamma)
```



# Implicit and explicit methods for reducing estimation bias
## Models handled by <tt>glm.fit</tt>


+ Observations on independent random variables $Y_1, \ldots, Y_n$ each
with density/mass function of the form
$f_{Y_i}(y) = \exp\left\{\frac{y \theta_i - b(\theta_i)-c(y)}{\phi/r_i} - a\left(-\frac{r_i}{\phi}\right) + m(y) \right\}$

# <tt>brglmFit</tt> method for <tt>glm</tt>
## Expansion of the MLE bias for GLMs
## Unifying iterative procedure
## <tt>enrich</tt> method for <tt>family</tt> objects

# Multinomial logistic regression
## Poisson trick and maximum likelihood
## Poisson trick and bias reduction

# Adjusted $z$-tests

# Discussion

``` {r, echo=FALSE, eval=FALSE}
browseURL(rmarkdown::render(system.file("inst", "useR!2016/kosmidis_useR!2016.Rmd", package = "brglm2")))
```

<!-- "fade", "default" "slide", "convex", "concave", "zoom" or "none" -->
<!-- “default”, “simple”, “sky”, “beige”, “serif”, “solarized”, “blood”, “moon”, “night”, “black”, “league” or “white” -->
