---
title: "Adjustacent category logit models using **brglm2**"
author: "[Ioannis Kosmidis](http://www.ucl.ac.uk/~ucakiko/)"
date: "05 February 2019"
output: rmarkdown::html_vignette
bibliography: brglm2.bib
nocite: |
  @kosmidis:2019
vignette: >
  %\VignetteIndexEntry{Adjustacent category logit models using brglm2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

# **bracl**
The [**brglm2**](https://github.com/ikosmidis/brglm2) R package
provides `bracl` which is a wrapper of `brglmFit` for fitting adjacent
category models for orginal responses using either maximum likelihood
or any of the various bias reduction methods described in
`brglmFit`. There is a formal equivalence between adjacent category
logit models for orginal responses and multinomial logistic regression
models (see, e.g. the [Multinomial logistic regression using
brglm2](https://cran.r-project.org/package=brglm2/vignettes/multinomial.html)
vignette and the `brmultinom` function). `bracl` utilised that
equivalence and fits the corresponding Poisson log-linear model, by
appropriately re-scaling the Poisson means to match the multinomial
totals (a.k.a. the "Poisson trick"). The mathematical details and
algorithm on using the Poisson trick for mean-bias reduction are given
in @kosmidis:11.



# Citation
If you found this vignette or **brglm2**, in general, useful, please consider citing **brglm2** and the associated paper. You can find information on how to do this by typing `citation("brglm2")`.

# Opinion on stem cell research and religious fundamentalism

The `stemcell` data set ships with **brglm2**. @agresti:15[, Section 4.1] provides a detailsed description of the variables recorded in this data set (see also `?stemcell`).
```{r, echo = TRUE}
library("brglm2")
data("stemcell", package = "brglm2")
```
## Maximum likelihood estimation
The following chunk of code fits an adjacent category logit model with proportional odds and reproduces @agresti:10[, Table 4.2]. Note that the intercept parameters are different because @agresti:10[, Table 4.2] uses different contrasts for the intercept parameters.
```{r, echo = TRUE}
stemcells_ml <- bracl(research ~ as.numeric(religion) + gender, weights = frequency, data = stemcell,
                      parallel = TRUE, type = "ML")
summary(stemcells_ml)
```
`stemcells_ml` is an object inheriting from
```{r, echo = TRUE}
class(stemcells_ml)
```
**brglm2** implements `print`, `coef`, `fitted`, `predict`, `summary`, `vcov` and `logLik` methods for


We can check if a model with non-proportional odds fits the data equally well by fitting it and carrying out a likelihood ration test.
```{r, echo = TRUE}
stemcells_ml_full <- bracl(research ~ as.numeric(religion) + gender, weights = frequency, data = stemcell,
                           parallel = FALSE, type = "ML")
summary(stemcells_ml_full)
```
The value of the log likelihood ratio statistic here is
```{r, echo = TRUE}
(lrt <- deviance(stemcells_ml) - deviance(stemcells_ml_full))
```
and has an asymptotic chi-squared distribution with
```{r, echo = TRUE}
(df1 <- df.residual(stemcells_ml) - df.residual(stemcells_ml_full))
```
The p-value from testing the hypothesis that `stemcells_ml_full` is an as good fit as `stemcells_ml` is
```{r, echo = TRUE}
pchisq(lrt, df1, lower.tail = FALSE)
```
hence, the simpler model is found to be as adequate as the full model is.

# Mean and median bias reduction
We can use `bracl` to fit the adjacent category model using estiamtors with smaller mean or median bias. For mean bias reduction we do
```{r, echo = TRUE}
summary(update(stemcells_ml, type = "AS_mean"))
```
and for median
```{r, echo = TRUE}
summary(update(stemcells_ml, type = "AS_median"))
```
The estimates from mean and median bias reduction are similar to the maximum likelihood ones, indicating that estimation bias is not a major issue here.

# Prediction
We can predict the category probabilities using the `predict` method
```{r, echo = TRUE}
predict(stemcells_ml, type = "probs")
```

# Relevant resources
`?brglmFit` and `?brglm_control` contain quick descriptions of the various bias reduction methods supported in **brglm2**. The [`iteration`](https://cran.r-project.org/package=brglm2/brglm2.pdf) vignette describes the iteration and gives the mathematical details for the bias-reducing adjustments to the score functions for generalized linear models.


# Citation
If you found this vignette or **brglm2**, in general, useful, please consider citing **brglm2** and the associated paper. You can find information on how to do this by typing `citation("brglm2")`.


# References
